---
title: "Supporting Information 1: R package usage"
format: pdf
editor: visual
fig-format: png
date: today
crossref: 
  fig-title: "Supplementary Figure 1."
  fig-prefix: "Sup. Figure 1."
  tbl-title: "Supplementary Table 1."
  tbl-prefix: "Sup. Table 1."
  chapters: true
bibliography: references.bib
---

This vignette contains the code and figures for comparing the relative usage of the `R` packages `gllvm` [@gllvm_off], `hmsc` [@hmsc_off], `boral` [@boral_off] and `ecoCopula` [@ecoCopula_off], which are all bespoke model-based ordination packages, with the `vegan` package [@vegan_off], a go-to for classical ordination, based on monthly downloads of each package from the Comprehensive R Archive Network (CRAN). In order to not confound comparison with other types of models, we did not include other general-purpose packages also capable of fitting GLLVMs, such as `glmmTMB` [@glmmTMB_off] and `VGAM` [@VGAM_off].

### Data setup

```{r}
#| output: false
# load required R packages
library(cranlogs) # get package download data from cran
library(dplyr) # data handling
library(lubridate) # aggregate data by date
library(ggplot2) # plotting
```

We begin by loading the data, from the beginning of 2020 to the end of 2025.

```{r}
#| output: false
#| warning: false
data_downloads <- cran_downloads(packages=c("vegan","gllvm","Hmsc","boral",
                                            "ecoCopula"),
                                 from = "2020-01-01", to = "2025-12-31")
downloads_monthly <- data_downloads |> 
  group_by(month=floor_date(date, "month"),package) |> 
  summarise(count=sum(count))
```

We also calculate the monthly ratio of downloads for each package:

```{r}
downloads_monthly <- downloads_monthly |> mutate(ratio=count/sum(count))
```

## Plotting

We then plot the data in a graph:

```{r}
#| echo: true
#| warning: false
#| label: fig-cit
#| fig-cap: "yearly citations for the vegan R package (classical ordination tools) and R packages in which model-based ordination tools are central (boral, gllvm and hmsc)."
#| fig-subcap: 
#| - "Total number of monthly citations for each package (log scale)"
#| - "Percentage of total citations across all packages per month"
#| fig-width: 9
#| fig-height: 4
ggplot(downloads_monthly) + # total downloads/month
  geom_line(aes(x=month, y=count, color=package),stat="identity") +
  scale_y_continuous(trans='log10') +
  theme_classic() +
  labs(color="Package", y="Total monthly downloads (log scale)", x="Date", 
       title="Monthly CRAN downloads (2020-2025)") +
  scale_color_brewer(palette="Set3") +
  scale_x_date(date_breaks = "1 year")
ggplot(downloads_monthly) +  # relative downloads/month
  geom_line(aes(x=month, y=ratio, color=package),stat="identity") +
  labs(color="Package", y="Relative monthly downloads", x="Date", 
       title="Relative number of monthly CRAN downloads (2020-2025)") +
  scale_color_brewer(palette="Set3") +
  theme_classic() +
  scale_x_date(date_breaks = "1 year")
```

\newpage

# References
